---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

lgr::get_logger("mlr3")$set_threshold("warn")
options(datatable.print.class = FALSE, datatable.print.keys = FALSE)
library(checkmate)
library(data.table)
library(mlr3misc)
```


# mlr3forecast

Extending mlr3 to time series forecasting.

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![RCMD Check](https://github.com/mlr-org/mlr3forecast/actions/workflows/rcmdcheck.yaml/badge.svg)](https://github.com/mlr-org/mlr3forecast/actions/workflows/rcmdcheck.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/mlr3forecast)](https://CRAN.R-project.org/package=mlr3forecast)
[![StackOverflow](https://img.shields.io/badge/stackoverflow-mlr3-orange.svg)](https://stackoverflow.com/questions/tagged/mlr3)
[![Mattermost](https://img.shields.io/badge/chat-mattermost-orange.svg)](https://lmmisld-lmu-stats-slds.srv.mwn.de/mlr_invite/)
<!-- badges: end -->

```{=gfm}
> [!IMPORTANT]
> This package is in an early stage of development and should be considered experimental.
> If you are interested in experimenting with it, we welcome your feedback!
```

## Installation

Install the development version from [GitHub](https://github.com/):

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("mlr-org/mlr3forecast")
```

## Usage

### Example: native forecast learner

First lets create a helper function to generate new data for forecasting tasks.

```{r}
library(mlr3forecast)

generate_newdata = function(task, n = 1L, resolution = "day") {
  assert_count(n)
  assert_string(resolution)
  assert_choice(
    resolution, c("second", "minute", "hour", "day", "week", "month", "quarter", "year")
  )

  order_cols = task$col_roles$order
  max_index = max(task$data(cols = order_cols)[[1L]])

  unit = switch(resolution,
    second = "sec",
    minute = "min",
    hour = ,
    day = ,
    week = ,
    month = ,
    quarter = ,
    year = identity(resolution),
    stopf("Invalid resolution")
  )
  unit = sprintf("1 %s", unit)
  index = seq(max_index, length.out = n + 1L, by = unit)
  index = index[2:length(index)]

  newdata = data.frame(index = index, target = rep(NA_real_, n), check.names = FALSE)
  setNames(newdata, c(order_cols, task$target_names))
}

task = tsk("airpassengers")
newdata = generate_newdata(task, 12L, "month")
newdata
```

Currently, we support native forecasting learners from the forecast package.
In the future, we plan to support more forecasting learners.

```{r}
task = tsk("airpassengers")
learner = lrn("fcst.arima", order = c(2L, 1L, 2L))$train(task)
prediction = learner$predict(task, 140:144)
prediction$score(msr("regr.rmse"))
newdata = generate_newdata(task, 12L, "month")
learner$predict_newdata(newdata, task)

learner = lrn("fcst.auto_arima")$train(task)
prediction = learner$predict(task, 140:144)
prediction$score(msr("regr.rmse"))
newdata = generate_newdata(task, 12L, "month")
learner$predict_newdata(newdata, task)

# works with quantile response
learner = lrn("fcst.auto_arima",
  predict_type = "quantiles",
  quantiles = c(0.1, 0.15, 0.5, 0.85, 0.9),
  quantile_response = 0.5
)$train(task)
learner$predict_newdata(newdata, task)
```

### machine learning forecasting

```{r, message = FALSE}
library(mlr3learners)

task = tsk("airpassengers")
task$select(setdiff(task$feature_names, "date"))
flrn = ForecastLearner$new(lrn("regr.ranger"), 1:12)$train(task)
newdata = data.frame(passengers = rep(NA_real_, 3L))
prediction = flrn$predict_newdata(newdata, task)
prediction
prediction = flrn$predict(task, 142:144)
prediction
prediction$score(msr("regr.rmse"))

flrn = ForecastLearner$new(lrn("regr.ranger"), 1:12)
resampling = rsmp("forecast_holdout", ratio = 0.9)
rr = resample(task, flrn, resampling)
rr$aggregate(msr("regr.rmse"))

resampling = rsmp("forecast_cv")
rr = resample(task, flrn, resampling)
rr$aggregate(msr("regr.rmse"))
```

### Multivariate

```{r}
library(mlr3pipelines)

task = tsk("airpassengers")
# datefeatures currently requires POSIXct
graph = ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(is_day = FALSE, hour = FALSE, minute = FALSE, second = FALSE)
  )
new_task = graph$train(task)[[1L]]
flrn = ForecastLearner$new(lrn("regr.ranger"), 1:12)$train(new_task)
prediction = flrn$predict(new_task, 142:144)
prediction$score(msr("regr.rmse"))

row_ids = new_task$nrow - 0:2
flrn$predict_newdata(new_task$data(rows = row_ids), new_task)
newdata = new_task$data(rows = row_ids, cols = new_task$feature_names)
flrn$predict_newdata(newdata, new_task)

resampling = rsmp("forecast_holdout", ratio = 0.9)
rr = resample(new_task, flrn, resampling)
rr$aggregate(msr("regr.rmse"))

resampling = rsmp("forecast_cv")
rr = resample(new_task, flrn, resampling)
rr$aggregate(msr("regr.rmse"))
```

### mlr3pipelines integration

```{r}
graph = ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(is_day = FALSE, hour = FALSE, minute = FALSE, second = FALSE)
  )
flrn = ForecastLearner$new(lrn("regr.ranger"), 1:12)
glrn = as_learner(graph %>>% flrn)$train(task)
prediction = glrn$predict(task, 142:144)
prediction$score(msr("regr.rmse"))
```

### Example: Forecasting electricity demand

```{r}
library(mlr3learners)
library(mlr3pipelines)

task = tsibbledata::vic_elec |>
  as.data.table() |>
  setnames(tolower) |>
  _[
    year(time) == 2014L,
    .(
      demand = sum(demand) / 1e3,
      temperature = max(temperature),
      holiday = any(holiday)
    ),
    by = date
  ] |>
  as_task_fcst(id = "vic_elec", target = "demand", order = "date")

graph = ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE, second = FALSE
    )
  )
flrn = ForecastLearner$new(lrn("regr.ranger"), 1:3)
glrn = as_learner(graph %>>% flrn)$train(task)

max_date = task$data()[.N, date]
newdata = data.frame(
  date = max_date + 1:14,
  demand = rep(NA_real_, 14L),
  temperature = 26,
  holiday = c(TRUE, rep(FALSE, 13L))
)
prediction = glrn$predict_newdata(newdata, task)
prediction
```

### Global Forecasting

```{r, message = FALSE}
library(mlr3learners)
library(mlr3pipelines)
library(tsibble) # needs not be loaded for it to somehow work

task = tsibbledata::aus_livestock |>
  as.data.table() |>
  setnames(tolower) |>
  _[, month := as.Date(month)] |>
  _[, .(count = sum(count)), by = .(state, month)] |>
  setorder(state, month) |>
  as_task_fcst(id = "aus_livestock", target = "count", order = "month", key = "state")

graph = ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      week_of_year = FALSE, day_of_week = FALSE, day_of_month = FALSE,
      day_of_year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE,
      second = FALSE
    )
  )
task = graph$train(task)[[1L]]
task$col_roles$key = "state"

flrn = ForecastLearner$new(lrn("regr.ranger"), 1:3)$train(task)
prediction = flrn$predict(task, 4460:4464)
prediction$score(msr("regr.rmse"))

flrn = ForecastLearner$new(lrn("regr.ranger"), 1:3)
resampling = rsmp("forecast_holdout", ratio = 0.9)
rr = resample(task, flrn, resampling)
rr$aggregate(msr("regr.rmse"))
```

### Example: Global vs Local Forecasting

In machine learning forecasting the difference between forecasting a time series
and longitudinal data is often refered to local and global forecasting.

```{r}
# TODO: find better task example, since the effect is minor here

graph = ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      week_of_year = FALSE, day_of_week = FALSE, day_of_month = FALSE,
      day_of_year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE,
      second = FALSE
    )
  )

# local forecasting
task = tsibbledata::aus_livestock |>
  as.data.table() |>
  setnames(tolower) |>
  _[, month := as.Date(month)] |>
  _[state == "Western Australia", .(count = sum(count)), by = .(month)] |>
  setorder(month) |>
  as_task_fcst(id = "aus_livestock", target = "count", order = "month")
task = graph$train(task)[[1L]]
flrn = ForecastLearner$new(lrn("regr.ranger"), 1L)$train(task)
tab = task$backend$data(
  rows = task$row_ids, cols = c(task$backend$primary_key, "month.year")
)
setnames(tab, c("row_id", "year"))
row_ids = tab[year >= 2015, row_id]
prediction = flrn$predict(task, row_ids)
prediction$score(msr("regr.rmse"))

# global forecasting
task = tsibbledata::aus_livestock |>
  as.data.table() |>
  setnames(tolower) |>
  _[, month := as.Date(month)] |>
  _[, .(count = sum(count)), by = .(state, month)] |>
  setorder(state, month) |>
  as_task_fcst(id = "aus_livestock", target = "count", order = "month", key = "state")
task = graph$train(task)[[1L]]
task$col_roles$key = "state"
flrn = ForecastLearner$new(lrn("regr.ranger"), 1L)$train(task)
tab = task$backend$data(
  rows = task$row_ids, cols = c(task$backend$primary_key, "month.year", "state")
)
setnames(tab, c("row_id", "year", "state"))
row_ids = tab[year >= 2015 & state == "Western Australia", row_id]
prediction = flrn$predict(task, row_ids)
prediction$score(msr("regr.rmse"))
```

### Example: Custom PipeOps

```{r}
library(mlr3learners)
library(mlr3pipelines)

task = tsk("airpassengers")
pop = po("fcst.lag", lag = 1:12)
new_task = pop$train(list(task))[[1L]]
new_task$data()

task = tsk("airpassengers")
graph = po("fcst.lag", lag = 1:12) %>>%
  ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      week_of_year = FALSE, day_of_week = FALSE, day_of_month = FALSE,
      day_of_year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE,
      second = FALSE
    )
  )
flrn = ForecastRecursiveLearner$new(lrn("regr.ranger"))
glrn = as_learner(graph %>>% flrn)$train(task)
prediction = glrn$predict(task, 142:144)
prediction$score(msr("regr.rmse"))

newdata = generate_newdata(task, 12L, "month")
glrn$predict_newdata(newdata, task)
```

### Example: common target transformations

Some common target transformations in forecasting are:

- differencing (WIP)
- log transformation, see example below
- power transformations such as [Box-Cox](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_boxcox.html) and [Yeo-Johnson](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_yeojohnson.html)
  currently only supported as feature transformation and not target
- scaling/normalization, available see [here](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_targettrafoscalerange.html)

```{r}
trafo = po("targetmutate",
  param_vals = list(
    trafo = function(x) log(x),
    inverter = function(x) list(response = exp(x$response))
  )
)

graph = po("fcst.lag", lag = 1:12) %>>%
  ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      week_of_year = FALSE, day_of_week = FALSE, day_of_month = FALSE,
      day_of_year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE,
      second = FALSE
    )
  )

task = tsk("airpassengers")
flrn = ForecastRecursiveLearner$new(lrn("regr.ranger"))
glrn = as_learner(graph %>>% flrn)
pipeline = ppl("targettrafo", graph = glrn, trafo_pipeop = trafo)
glrn = as_learner(pipeline)$train(task)
prediction = glrn$predict(task, 142:144)
prediction$score(msr("regr.rmse"))
```

```{r, eval = FALSE}
graph = po("fcst.lag", lag = 1:12) %>>%
  ppl("convert_types", "Date", "POSIXct") %>>%
  po("datefeatures",
    param_vals = list(
      week_of_year = FALSE, day_of_week = FALSE, day_of_month = FALSE,
      day_of_year = FALSE, is_day = FALSE, hour = FALSE, minute = FALSE,
      second = FALSE
    )
  )

task = tsk("airpassengers")
flrn = ForecastRecursiveLearner$new(lrn("regr.ranger"))
glrn = as_learner(graph %>>% flrn)
trafo = po("fcst.targetdiff", lag = 12L)
pipeline = ppl("targettrafo", graph = glrn, trafo_pipeop = trafo)
glrn = as_learner(pipeline)$train(task)
prediction = glrn$predict(task, 142:144)
prediction$score(msr("regr.rmse"))
```
